Bayesian Rate Estimation

### 2012-08-09 many.tex
Added comments on Section II.

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

@@

@@

\author{Ilya Mandel}
\email{ilya@chgk.info}
\homepage{http://www.chgk.info/~ilyamandel/}
\affiliation{School of Physics and Astronomy\\University of Birmingham\\Edgbaston B15 2TT Birmingham\\United Kingdom}
%\affiliation{University of Birmingham}

@@

\begin{abstract}
  We show how to obtain a Bayesian estimate of the rate of signal
  events from a set of signal and background events when the shapes of
  the signal and background distributions are known, can be estimated,
  or approximated; our method works well even if the foreground and
  background event distributions overlap significantly.  We give
  examples of determining the rates of gravitaitonal wave events in
  the presence of background triggers from a template bank when noise
  parameters are known and/or can be fit from the trigger data.  We
  also give an example of determining globular cluster shape and
  location parameters from an observation of a stellar field that
  contains a non-uniform background density of stars superimposed on
  the cluster stars.
\end{abstract}

@@

We assume that both the foreground and background events are samples
from an inhomogeneous Poisson process with differential rates
\begin{equation}
  \frac{dN_f}{dx} = f(x, \theta)
\end{equation}
and 
\begin{equation}
  \frac{dN_b}{dx} = b(x, \theta),
\end{equation}
where the $\theta$ argument represents additional ``shape'' parameters
that may affect the distribution, and for which we will eventually
fit.  The cumulative rates of the two processes are therefore
\begin{equation}
  F(x,\theta) \equiv \int_{-\infty}^x ds\, f(s, \theta)
\end{equation}
and
\begin{equation}
  B(x,\theta) \equiv \int_{-\infty}^x ds\, b(s,\theta).
\end{equation}
The assumption that the foreground and background events form an
inhomogeneous Poisson process implies
\begin{enumerate}
\item \label{prop:Poisson} The number of events in any range of
  ranking statistics, $x \in [x_1, x_2]$ is Poisson distributed with
  rate $F(x_2, \theta) - F(x_1,\theta)$ or $B(x_2,\theta) -
  B(x_1,\theta)$.
\item The numbers of events in non-overlapping ranges of ranking
  statistics are independent. 
\item The probability of exactly one foreground event between $x$ and
  $x+h$ is given by
  \begin{equation}
    P(N = 1 \in [x, x+h]) = f(x,\theta) h + \order{h^2}.
  \end{equation}
  and similarly for background events.
\item The probability of two or more events in a small range of
  ranking statistic is negligable
  \begin{equation}
    P(N = 2 \in [x, x+h]) = \order{h^2}.
  \end{equation}
\end{enumerate}
The foreground and background rates can in general depend on several
parameters; the goal of our analysis is to determine the posterior
probability distributions for these parameters that are implied by the
data.  At the least, we will want to know the overall amplitude of the
foreground and background rates.  Let
\begin{equation}
  f(x,\theta) = R_f \hat{f}(x,\theta'),
\end{equation}
and 
\begin{equation}
  b(x, \theta) = R_b \hat{b}(x, \theta'),
\end{equation}
where $\hat{F}(\infty, \theta') = \hat{B}(\infty, \theta') = 1$, and
$\theta' = \theta \setminus \{R_{f}, R_{b} \}$.  Then $R_f$ and $R_b$
are the total number of foreground and background events expected and
$\hat{f}(x, \theta')$ and $\hat{b}(x, \theta')$ are the likelihood of
obtaining an event with ranking statistic $x$ under the foreground and
background distributions.  In what follows, we will drop the prime,
using $\theta$ to denote all parameters of the rate distributions
except $R_f$ and $R_b$.

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $f_i$, which is
either 0 or 1, indicating a background or foreground event.  These
``state'' flags are parameters in our model, along with $R_f$, $R_b$,
and $\theta$.  We can marginalize over our uncertainty in the state of
any given event by summing posteriors over $f_i = \mathset{0,1}$.

Bayes' theorem relates the posterior probability of the state flags,
rates, and shape parameters, $p\left(\mathset{f_i}, R_f, R_b, \theta |
  d \right)$, the likelihood of the data, $p\left( d | \mathset{f_i},
  R_f, R_b, \theta\right)$, and the prior probability of state flags,
rates and shape parameters before any data are obtained, $p\left(
  \mathset{f_i}, R_f, R_b, \theta\right)$:
\begin{equation}
  \label{eq:Bayes}
  p\left(\mathset{f_i}, R_f, R_b, \theta | d \right)  \propto
  p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)
  p\left(\mathset{f_i}, R_f, R_b, \theta\right). 
\end{equation}
The constant of proportionality, called the evidence, $p(d)$, is a
normalization constant independent of the state flags, rates, and
shape parameters.

Each foreground event is drawn from the probability distribution
$\hat{f}$ and each background event is drawn from the probability
distribution $\hat{b}$.  The events are independent of each other.
Therefore, the likelihood of the data is 
\begin{equation}
  p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)  = \left[
    \prod_{\mathset{i | f_i = 1}} \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} \hat{b}\left( x_i, \theta\right)
  \right].  
\end{equation}
The prior distribution can be factorized as
\begin{equation}
  \label{eq:combined-flag-rate-prior}
  p\left(\mathset{f_i}, R_f, R_b, \theta\right)  = p\left(
    \mathset{f_i} | R_f, R_b, \theta\right)p(R_f, R_b, \theta).
\end{equation}
The first term is not acutally a ``prior'' in the usual sense; the
conditional probability of the $\mathset{f_i}$ on the rates is given
by
\begin{equation}
  \label{eq:flag-conditional-prior}
  p\left(\mathset{f_i} | R_f, R_b, \theta\right) = \frac{R_f^{N_f}
    R_b^{N_b}}{N_f! N_b!} \exp\left[ - \left(R_f + R_b\right) \right]
  N_f! N_b! = R_f^{N_f}R_b^{N_b} \exp\left[ - \left(R_f + R_b\right) \right],
\end{equation}
where $N_f$ and $N_b$ are the number of foreground and background
events.  This expression follows from Property \ref{prop:Poisson} of
inhomogenous Poisson processes and the $N_f! N_b!$ differnt ways of
assigning foreground and background flags to events for fixed $N_f$,
$N_b$.

The second term in Eq.~\eqref{eq:combined-flag-rate-prior} is a
traditional prior.  Because the rate parameters enter the posterior in
the same form as Poisson rates, we choose here the Poisson Jeffrey's
prior on rates, independent of the shape parameters
\begin{equation}
  p\left( R_f, R_b, \theta\right) = \frac{1}{\sqrt{R_f R_b}} p(\theta),
\end{equation}
but of course other choices are possible.  This choice has the
advantage that the prior is normalizable as $R_f, R_b \to 0$, and the
exponentials in Eq.~\eqref{eq:flag-conditional-prior} regularize the
posterior as $R_f, R_b \to \infty$.

Putting everything together, the posterior is
\begin{equation}
  \label{eq:posterior}
  p\left( \mathset{f_i}, R_f, R_b, \theta | d \right) \propto \left[
    \prod_{\mathset{i | f_i = 1}} R_f \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} R_b \hat{b}\left( x_i, \theta\right)
  \right] \exp\left[ - \left(R_f + R_b\right) \right]
  \frac{p(\theta)}{\sqrt{R_f R_b}}
\end{equation}
Not surprisingly, this posterior equivalent to one derived by assuming
that the flags, $\mathset{f_i}$, are un-observed data and treating the
sets $\mathset{x_i | f_i = 1}$ and $\mathset{x_i | f_i = 0}$ as
samples from an inhomogeneous Poisson process:
\begin{multline}
  p\left( \mathset{f_i}, R_f, R_b, \theta | d \right) \propto p\left(
    d, \mathset{f_i} | R_f, R_b, \theta \right) p\left( R_f, R_b,
    \theta \right) \\ = \left[ \prod_{\mathset{i|f_i=1}} f\left( x_i,
      \theta \right) \right] \left[ \prod_{\mathset{i|f_i=0}} b\left(
      x_i, \theta\right) \right] \exp\left[ -\left( R_f + R_b\right)
  \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}

@@

Eq.~\eqref{eq:posterior} is unchanged if the ranking statistic is
multi-dimensional; in this case, the rates are 
\begin{equation}
  R_f = \int d^n x \, f(x, \theta)
\end{equation}
and
\begin{equation}
  R_b = \int d^n x \, b(x, \theta),  
\end{equation}
where $f$ and $b$ are rate densities on the $n$-dimensional space of
ranking statistics.  We give an example of fitting for
multi-dimensional rate densities in \S~\ref{sec:star-cluster}.

>>>>

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color}

@@

\newcommand{\ilya}[1]{{\color{red} \bf #1}}
\newcommand{\will}[1]{{\color{blue} \bf #1}}
\newcommand{\jon}[1]{{\color{orange} \bf  #1}}

@@

\author{Ilya Mandel}
\email{imandel@star.sr.bham.ac.uk}
\homepage{http://www.star.sr.bham.ac.uk/~imandel}
\affiliation{School of Physics and Astronomy\\University of Birmingham\\Edgbaston B15 2TT Birmingham\\United Kingdom}
%\affiliation{University of Birmingham}

@@

\begin{abstract}
  We show how to obtain a Bayesian estimate of the rate of signal
  events from a set of signal and background events when the shapes of
  the signal and background distributions are known, can be estimated,
  or approximated; our method works well even if the foreground and
  background event distributions overlap significantly.  We give
  examples of determining the rates of gravitatonal-wave events in
  the presence of background triggers from a template bank when noise
  parameters are known and/or can be fit from the trigger data.  We
  also give an example of determining globular-cluster shape and
  location parameters from an observation of a stellar field that
  contains a non-uniform background density of stars superimposed on
  the cluster stars.
\end{abstract}

@@

We assume that both the foreground and background events are samples
from an inhomogeneous \ilya{[pardon my ignorance, what does "inhomogeneous" refer to in this context (i.e., what would a homogeneous Poisson process be?)]} Poisson process with \ilya{respective} differential rates
\begin{equation}
  \frac{dN_f}{dx} = f(x, \theta)
\end{equation}
and 
\begin{equation}
  \frac{dN_b}{dx} = b(x, \theta),
\end{equation}
where the $\theta$ argument represents additional ``shape'' parameters
that may affect the distribution, and for which we will eventually
fit.  The cumulative rates of the two processes are therefore
\begin{equation}
  F(x,\theta) \equiv \int_{-\infty}^x ds\, f(s, \theta)
\end{equation}
and
\begin{equation}
  B(x,\theta) \equiv \int_{-\infty}^x ds\, b(s,\theta).
\end{equation}
The assumption that the foreground and background events form an
inhomogeneous Poisson process implies
\begin{enumerate}
\item \label{prop:Poisson} The number of events in any range of
  ranking statistics, $x \in [x_1, x_2]$ is Poisson distributed with
  rate $F(x_2, \theta) - F(x_1,\theta)$ or $B(x_2,\theta) -
  B(x_1,\theta)$.
\item The numbers of events in non-overlapping ranges of ranking
  statistics are independent. 
\item The probability of exactly one foreground event between $x$ and
  $x+h$ is given by
  \begin{equation}
    P(\ilya{n} = 1 \in [x, x+h]) = f(x,\theta) h + \order{h^2}.
  \end{equation}
  and similarly for background events.
\item The probability of two or more events in a small range of
  ranking statistic is negligible
  \begin{equation}
    P(\ilya{n} = 2 \in [x, x+h]) = \order{h^2}.
  \end{equation}
\end{enumerate}
The foreground and background rates can in general depend on several
parameters; the goal of our analysis is to determine the posterior
probability distributions for these parameters that are implied by the
data.  At the least, we will want to know the overall amplitude of the
foreground and background rates.  Let
\begin{equation}
  f(x,\theta) = R_f \hat{f}(x,\theta'),
\end{equation}
and 
\begin{equation}
  b(x, \theta) = R_b \hat{b}(x, \theta'),
\end{equation}
where $\hat{F}(\infty, \theta') = \hat{B}(\infty, \theta') = 1$, and
$\theta' = \theta \setminus \{R_{f}, R_{b} \}$.  Then $R_f\ilya{\equiv F(\infty,\theta)}$ and $R_b\ilya{\equiv B(\infty,\theta)}$
are the total number of foreground and background events expected and
$\hat{f}(x, \theta')$ and $\hat{b}(x, \theta')$ are the likelihood of
obtaining an event with ranking statistic $x$ under the foreground and
background distributions.  In what follows, we will drop the prime,
using $\theta$ to denote all parameters of the rate distributions
except $R_f$ and $R_b$.

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $f_i$, which is
either 0 \ilya{(background)} or 1 \ilya{(foreground)}.  These ``state'' flags are parameters in our model, along with $R_f$, $R_b$, and $\theta$.  We can marginalize over our uncertainty in the state of
any given event by summing posteriors over $f_i = \mathset{0,1}$.

Bayes' theorem relates the posterior probability of the state flags,
rates, and shape parameters, $p\left(\mathset{f_i}, R_f, R_b, \theta |
  d \right)$, the likelihood of the data, $p\left( d | \mathset{f_i},
  R_f, R_b, \theta\right)$, and the prior probability of state flags,
rates and shape parameters before any data are obtained, $p\left(
  \mathset{f_i}, R_f, R_b, \theta\right)$:
\begin{equation}
  \label{eq:Bayes}
  p\left(\mathset{f_i}, R_f, R_b, \theta | d \right)  \ilya{=}
  \frac{p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)
  p\left(\mathset{f_i}, R_f, R_b, \theta\right)}{\ilya{p(d)}}. 
\end{equation}
The \ilya{normalization constant}, called the evidence, $p(d)$, is independent of the state flags, rates, and shape parameters.

Each foreground event is drawn from the probability distribution
$\hat{f}$ and each background event is drawn from the probability
distribution $\hat{b}$.  The events are independent of each other.
Therefore, the likelihood of the data is 
\begin{equation}
  p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)  = \left[
    \prod_{\mathset{i | f_i = 1}} \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} \hat{b}\left( x_i, \theta\right)
  \right].  
\end{equation}
The prior distribution can be factorized as
\begin{equation}
  \label{eq:combined-flag-rate-prior}
  p\left(\mathset{f_i}, R_f, R_b, \theta\right)  = p\left(
    \mathset{f_i} | R_f, R_b, \theta\right)p(R_f, R_b, \theta).
\end{equation}
The first term is not actually a ``prior'' in the usual sense; the
conditional probability of the flags $\mathset{f_i}$ on the rates is given
by
\begin{equation}
  \label{eq:flag-conditional-prior}
  p\left(\mathset{f_i} | R_f, R_b, \theta\right) = \frac{R_f^{N_f}
    R_b^{N_b}}{N_f! N_b!} \exp\left[ - \left(R_f + R_b\right) \right]
  N_f! N_b! = R_f^{N_f}R_b^{N_b} \exp\left[ - \left(R_f + R_b\right) \right],
\end{equation}
where $N_f$ and $N_b$ are the number of foreground and background
events.  This expression follows from Property \ref{prop:Poisson} of
inhomogeneous Poisson processes and the $N_f! N_b!$ different ways of
assigning foreground and background flags to events for fixed $N_f$,
$N_b$.

\ilya{[I agree with the final result, but I find this way of deriving it to be confusing.  First of all, the number of ways to assign flags to events for fixed $N_f$ and $N_b$ is $N!/(N_f! N_b!)$, not $N_f! N_b!$ as written.  Secondly, you are computing the probability of having a particular ordered list of flags, so there's no reason to consider different permutations.  Thirdly, you are not properly including the fact that the length of the flags vector must be exactly $N$, the total number of data points.  (Well, you have the right result, so you are obviously doing these things properly, but the explanation just doesn't seem clear to me. :) )  I would make that $N$ explicit, by writing Eq.~(12) as

$$
  p\left(\mathset{f_i}, N, R_f, R_b, \theta\right)  = 
  p\left( \mathset{f_i} | N, R_f, R_b\right) p\left(N |R_f, R_b\right)  
 p\left(R_f, R_b, \theta\right).
$$

Then 
$$ p\left( \mathset{f_i} | N, R_f, R_b\right) = 
\prod_{\mathset{i|f_i=1}} \left(\frac{R_f}{R_f+R_b}\right) 
\prod_{\mathset{i|f_i=0}} \left(\frac{R_b}{R_f+R_b}\right) =
\left(\frac{R_f}{R_f+R_b}\right)^{N_f} \left(\frac{R_b}{R_f+R_b}\right)^{N_b}. 
$$
Note that there are no permutations involved here -- once we assume that there are precisely $N=N_f+N_b$ events, the probability for the i'th event to have flag $f_i$ is either $R_f/(R_f+R_b)$ or $R_b/(R_f+R_b)$ depending on whether $f_i$ is 1 or 0.
Meanwhile,
$$
p\left(N |R_f, R_b\right) = \frac{\left(R_f+R_b\right)^N}{N!} e^{-(R_f+R_b)},
$$
since the distribution of total event number is a Poisson process with rate $R_f+R_b$.
Multiplying these, we get the same result as in (13), but in a way that even I can understand. :)
]}

The second term in Eq.~\eqref{eq:combined-flag-rate-prior} is a
traditional prior.  Because the rate parameters enter the posterior in
the same form as Poisson rates, we choose here the Poisson Jeffrey's
prior on rates \ilya{[citation?]}, independent of the shape parameters
\begin{equation}
  p\left( R_f, R_b, \theta\right) \ilya{\propto} \frac{1}{\sqrt{R_f R_b}} p(\theta),
\end{equation}
but of course other choices are possible.  This choice has the
advantage that the prior is normalizable as $R_f, R_b \to 0$, and the
exponentials in Eq.~\eqref{eq:flag-conditional-prior} regularize the
posterior as $R_f, R_b \to \infty$.

Putting everything together, the posterior is
\begin{equation}
  \label{eq:posterior}
  p\left( \mathset{f_i}, R_f, R_b, \theta | d \right) \propto \left[
    \prod_{\mathset{i | f_i = 1}} R_f \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} R_b \hat{b}\left( x_i, \theta\right)
  \right] \exp\left[ - \left(R_f + R_b\right) \right]
  \frac{p(\theta)}{\sqrt{R_f R_b}}
\end{equation}
Not surprisingly, this posterior is equivalent to one derived by assuming
that the flags, $\mathset{f_i}$, are un-observed data and treating the
sets $\mathset{x_i | f_i = 1}$ and $\mathset{x_i | f_i = 0}$ as
samples from an inhomogeneous Poisson process:
\begin{multline}
  p\left( \mathset{f_i}, R_f, R_b, \theta | d \right) \propto p\left(
    d, \mathset{f_i} | R_f, R_b, \theta \right) p\left( R_f, R_b,
    \theta \right) \\ = \left[ \prod_{\mathset{i|f_i=1}} f\left( x_i,
      \theta \right) \right] \left[ \prod_{\mathset{i|f_i=0}} b\left(
      x_i, \theta\right) \right] \exp\left[ -\left( R_f + R_b\right)
  \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}

\ilya{[I can't derive the preceding equation in 1 line. ;)  So I'd suggest either making the alternative derivation more explicit, or, if you don't think it's worth the space, just saying that one could derive the same thing in this alternative way -- otherwise, just repeating Eq.~(15) doesn't seem very useful.]}

@@

Eq.~\eqref{eq:posterior} is unchanged if the ranking statistic is
multi-dimensional; in this case, the rates are 
\begin{equation}
  R_f = \int d^\ilya{k} \ilya{\vec{x}} \, f(x, \theta)
\end{equation}
and
\begin{equation}
  R_b = \int d^\ilya{k} \ilya{\vec{x}} \, b(x, \theta),  
\end{equation}
where $f$ and $b$ are rate densities on the \ilya{$k$}-dimensional space of
ranking statistics.  We give an example of fitting for
multi-dimensional rate densities in \S~\ref{sec:star-cluster}.
